The impacts of park access on health outcomes: A spatial comprehensive approach

### 2019-02-22 manuscript.Rnw
Revise literature review and physical activity model discussion

In spite of this interest and the numerous studies attempting to empirically
estimate the strength of such support, evidence to date has been mixed.
A major challenge researchers have faced is a proliferation of techniques for
measuring access to green spaces, of which many are plausible but lack
analytical or theoretical basis. In this paper, we present a holistic
and flexible measurement for park accessibility based in random utility
choice theory. This choice-based accessibility measure compares the continuous
distance to all parks in the region, weighted against the size of the park and
its other amenities. We apply this measurement to study the link between park
accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatial correlation and tract-level socioeconomic characteristics.
We find a positive relationship where the least park-accessible tracts have an
expected physical activity participation rate half a percentage point lower than
the most accessible tracts, though this result is only suggestive of correlation
($p$-value $<0.1$ from a statistical) perspective.
However, we also find a strong and significant correlation between
increased park accessibility and decreased obesity rates, again controlling for
spatial correlation, socioeconomic attributes, and physical activity rates.

We also demonstrate how the choice-based accessibility metric could potentially
be expanded to account for other elements of a park's attractiveness by
introducing the number of geolocated tweets emanating from within each park during
September 2014. The paper concludes with a discussion of
opportunities for future research.

@@

# determine which tracts are neighbors of each other. Need to be less than
# 1.8 miles away from the centroids
tracts.dnn <- dnearneigh(as(pop_weighted_centroids, "Spatial"), 0, 1.8 * 5280)
# calculate the inverse distance to use as a weight on the neighbor connection
dists <- nbdists(tracts.dnn, as(pop_weighted_centroids, "Spatial"))
dists.inv <- lapply(dists, function(x) 1 / x)
# get the spatial weights matrix and its trace to use in any impact analyses
W <- nb2listw(neighbours = tracts.dnn, glist = dists.inv,
              zero.policy = TRUE, style = "W")
trMC <- trW(as(W, "CsparseMatrix"), type="MC") # trace used in montecarlo impacts
@
This is a similar specification to that used by  \citet{tra_autoregression_paper}
in a study of home prices in Atlanta. We selected this specification after
comparing overall model likelihoods between it and an adjacent (queen)
tract specification.

@@

We initially estimated potential values of the $\lambda$ coefficients by
applying a limited-memory, box-constrained optimization algorithm 
maximizing the model log-likehood.\footnote{Using the {\texttt nlm} function
in R \citep{RCoreTeam2018}} The constraint required that the coefficient on
size be positive and that on distance negative; all else equal, we assert people
will prefer to use larger parks and parks that are nearer to their residence. This
resulted in $\lambda_d = -1.98$  and $\lambda_s = 0.000$. Though the 
``most likely''
parameter value, in this case it was unsatisfactory for two reasons: first,
our prior assertion that park size is a valuable attribute of a park, and second
the observation that with these maximum likelihood values tracts bordering on 
the corners of
Central Park were measured with poor overall accessibility, a unacceptably 
unintuitive result. We subsequently 
adjusted the $\lambda_s$ value to $0.3$; the relative ratio of the coefficients 
implies that all else equal, an individual would travel 6.6 times further to
access a park twice as large.

<<logsum_betas>>=
# This is where we actually calculate the accessibility statistic, using the
# distances matrix between tracts (in the rows) and parks (in the columns).
tracts <- tracts %>%
  mutate(
    # Log-sum based utilities, without and with Twitter data
    access_ls = calculate_park_logsums(
      log(distances), parks$log_size, betas = c(-1.98, 0.3)),
    tweets_ls = calculate_park_logsums(
      log(distances), parks$log_size, parks$log_tweets,
      betas = c(-1.98, 0.3, 0.1)),
    # Normalize the logsum onto the Z-scale
    access_ls = (access_ls - mean(access_ls) ) / sd(access_ls),
    tweets_ls = (tweets_ls - mean(tweets_ls) ) / sd(tweets_ls)
  )

@@

We estimated SDM and SEM models regressing the physical activity rate against
against the base covariates (without the accessibility logsum). A likelihood
ratio test failed to reject that the SEM is equivalent
($\chi^2 = \Sexpr{test_sdmsem$Chisq[2]}$ on \Sexpr{test_sdmsem$Df[2]} degrees of
freedom), so we use the SEM specification only going forward.
Table \ref{tab:pa_models} presents the estimated coefficients for the base
model with income represented in two ways. In the first, the share of the tract
population in
the bottom three income deciles and the share of the population in the top three
income deciles are each included as independent predictors. In the second, the
logarithm of the mean tract income is included as a predictor.
Notwithstanding the higher model likelihood of the deciles model, comparing
coefficients across the models shows that the predictors change magnitude
or significance only marginally, if at all. Overall, the base model
predictors are significant and of intuitive direction. Tracts with higher shares
of full-time workers, college-educated adults, young adults, and high-income
households all show a greater share of individuals engaging in regular physical
activity. Conversely, tracts with greater population density and a greater
share of single adults, children, seniors, minorities of all types, and low
income households all report lower rates of physical activity.

The other two models presented in Table \ref{tab:pa_models} include
the log-sum accessibility statistic described in above added as a predictor. 
In both income specifications,
the access logsum is positively correlated with physical activity rates; all else
equal, people in tracts with higher accessibility to parks report higher rates of
physical activity. The estimated value of this coefficient and its standard
error is effectively the same with both representations of income, though in the
log representation the estimated $t$-statistic technically passes the $0.1$ level.
This is suggestive evidence only from a statistical perspective, but it deserves
comment. Taking a compromise value of $0.10$ between the two models, the data
suggest going from the least-accessible ($-3.0$) to the most-accessible ($3.0$) 
tract implies a $0.6$ percentage-point rise in physical activity rates, though
with a wide confidence interval around that estimate.

@@

<<screen_obese_models, eval = FALSE>>=
screenreg(obesity_models, digits = 4, table = FALSE,
          stars = c(0.001, 0.01, 0.05, 0.1) )
@

>>>>

In spite of this interest and the numerous studies attempting to empirically
estimate the strength of such support, evidence to date has been mixed.
A major challenge researchers have faced is a proliferation of techniques for
measuring access to green spaces, of which many are plausible but lack
analytical or theoretical basis. In this paper, we present a holistic
and flexible measurement for park accessibility based in random utility
choice theory. This choice-based accessibility measure compares the continuous
distance to all parks in the region, weighted against the size of the park and
its other amenities. We apply this measurement to study the link between park
accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatial correlation and tract-level socioeconomic characteristics.
We find a positive relationship where the least park-accessible tracts have an
expected physical activity participation rate roughly a full percentage point
lower than the most accessible tracts; a traditional quarter-mile buffer
analysis estimates an effect twice as large but marginally less significant.
However, we also find a strong and significant correlation between
the choice-based park accessibility metric and decreased obesity rates --- again
controlling for spatial correlation, socioeconomic attributes, and physical
activity rates. A traditional buffer analysis does not show significance.

We also demonstrate how the choice-based accessibility metric could potentially
be expanded to account for other measures of a park's attractiveness by
introducing the number of geolocated tweets emanating from within each park
during September 2014. The paper concludes with a discussion of opportunities
for future research.

@@

# determine which tracts are neighbors of each other. Need to be less than
# 1.8 miles away from the centroids
tracts.dnn <- dnearneigh(as(pop_weighted_centroids, "Spatial"), 0, 1.8 * 5280)
# calculate the inverse distance to use as a weight on the neighbor connection
dists <- nbdists(tracts.dnn, as(pop_weighted_centroids, "Spatial"))
dists.inv <- lapply(dists, function(x) 1 / x)
# get the spatial weights matrix and its trace to use in any impact analyses
W <- nb2listw(neighbours = tracts.dnn, glist = dists.inv,
              zero.policy = TRUE, style = "W")
trMC <- trW(as(W, "CsparseMatrix"), type="MC") # trace used in montecarlo impacts
@
This is a similar specification to that used by  \citet{tra_autoregression_paper}
in a study of home prices in Atlanta. We selected this specification after
comparing overall model likelihoods between it and specification where tracts
sharing a boundary are considered neighbors (queen-adjacent).

@@

We initially estimated potential values of the $\lambda$ coefficients by
applying a limited-memory, box-constrained optimization algorithm
maximizing the model log-likehood.\footnote{Using the {\texttt nlm} function
in R \citep{RCoreTeam2018}} The constraint required that the coefficient on
size be positive and that on distance negative; all else equal, we assert people
will prefer to use larger parks and parks that are nearer to their residence. This
resulted in $\lambda_d = -1.98$  and $\lambda_s = 0.000$. Though the ``most
likely'' parameter value, in this case it was unsatisfactory for two reasons:
first, our prior assertion that park size is a valuable attribute of a park, and
second the observation that with these maximum likelihood values tracts
bordering on the corners of Central Park were measured with poor overall
accessibility, an unacceptably unintuitive result. We subsequently adjusted the
$\lambda_s$ value to $0.3$; the relative ratio of the coefficients implies that
all else equal, an individual would travel 6.6 times further to access a park
twice as large.

<<logsum_betas, warning = FALSE>>=
# This is where we actually calculate the accessibility statistic, using the
# distances matrix between tracts (in the rows) and parks (in the columns).
tracts <- tracts %>%
  mutate(
    # Log-sum based utilities, without and with Twitter data
    access_ls = calculate_park_logsums(
      log(distances), parks$log_size, betas = c(-1.98, 0.3)),
    tweets_ls = calculate_park_logsums(
      log(distances), parks$log_size, parks$log_tweets,
      betas = c(-1.98, 0.3, 0.1)),
    # Normalize the logsum onto the Z-scale
    access_ls = (access_ls - mean(access_ls) ) / sd(access_ls),
    tweets_ls = (tweets_ls - mean(tweets_ls) ) / sd(tweets_ls)
  )

@@

We estimated SDM and SEM models regressing the physical activity rate against
against the base covariates (without the accessibility logsum). A likelihood
ratio test failed to reject that the SDM is different from the SEM, so we use
the SEM specification only going forward. Table \ref{tab:pa_models} presents the
estimated coefficients for the base model with the socioecononomic controls and
no representation of park accessibility. For the most part the coefficients are
highly significant and of the expected sign. Tracts with higher shares
of full-time workers, college-educated adults, young adults, and high-income
households all show a greater share of individuals engaging in regular physical
activity. Conversely, tracts with greater population density and a greater
share of single adults, children, seniors, minorities of all types, and low
income households all have lower modeled rates of physical activity.

To this base model we add a measure of park accessibility considered in two
ways: a logsum-based statistic as described earlier and a more typical quarter-mile
binary access buffer. The estimated coefficients for these models are also
given in Table \ref{tab:pa_models}. In both accessibility specifications,
the access measure is positively correlated with physical activity rates; all else
equal, people in tracts with higher accessibility to parks have higher estimated
rates of physical activity. Though the estimated coefficient on the quarter-mile
buffer appears to have a larger absolute value than the logsum-based statistic,
this masks a key difference in the specification worth remembering.
The logsum allow for variation in access unavailable in the buffer-based method.
In our dataset almost every tract
(\Sexpr{round(sum(tracts$access_025) / nrow(tracts) * 100, 0)} \%) is within
a quarter-mile of {\em some} kind of park, but this park may not be large enough
to support physical activities all individuals would like to participate in.
Because tracts with only adequate park access are grouped with tracts having
excellent park access, the overall effect is comparatively unclear. The logsum-based
coefficient, on the other hand, suggest going from the least-accessible ($-3.0$)
to the most-accessible ($3.0$) tract implies roughly a $1.1$\% percentage-point
rise in estimated physical activity rates

@@

<<screen_obese_models, eval = FALSE>>=
screenreg(obesity_models, digits = 4, table = FALSE,
          stars = c(0.001, 0.01, 0.05) )
@
